Figures_MSA-PD

### 2025-11-18 Manuscript_figures.Rmd
Update for resubmission

# Extended Data Figure 4

@@

```{r, fig.width=20, fig.height=8}
cowplot::plot_grid(plotlist = list(
  cao_pd$plotSampleDistances(space = "expression.shifts", show.sample.size = T, method = "UMAP", title = "Condition"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$age %>% setNames(rownames(meta)), title = "Age"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$brain_bank %>% setNames(rownames(meta)), show.sample.size = T, title = "Brain bank"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$extract %>% setNames(rownames(meta)), show.sample.size = T, title = "Extract batch"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$flowcell %>% setNames(rownames(meta)), show.sample.size = T, title = "Flowcell"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$pmi %>% setNames(rownames(meta)), title = "PMI"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$sex %>% setNames(rownames(meta)), show.sample.size = T, title = "Sex"),
  cao_pd$test.results$metadata.separation$pvalues %>%
    data.frame(p = .) %>%
    mutate(pp = -log10(p)) %>%
    tibble::rownames_to_column("covariate") %>%
    ggplot(aes(covariate, pp)) +
    geom_bar(stat = "identity", fill = "lightblue4") +
    ylim(0, 2) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    labs(x = "", y = "-log10(adj. p)", title = "Association significance") +
    geom_hline(yintercept = -log10(0.05), colour = "red3")
), ncol = 4) -> p

@@

```{r, fig.width=25, fig.height=8}
cowplot::plot_grid(plotlist = list(
  cao_msa$plotSampleDistances(space = "expression.shifts", show.sample.size = T, method = "UMAP", title = "Condition"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$age %>% setNames(rownames(meta)), title = "Age"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$disease_duration %>% setNames(rownames(meta)), show.sample.size = T, title = "Disease duration"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$extract %>% setNames(rownames(meta)), show.sample.size = T, title = "Extract batch"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$flowcell %>% setNames(rownames(meta)), show.sample.size = T, title = "Flowcell"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$pmi %>% setNames(rownames(meta)), title = "PMI"),
    cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$sex %>% setNames(rownames(meta)), show.sample.size = T, title = "Sex"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$subtype %>% setNames(rownames(meta)), show.sample.size = T, title = "Subtype"),
  cao_msa$test.results$metadata.separation$pvalues %>%
    data.frame(p = .) %>%
    mutate(pp = -log10(p)) %>%
    tibble::rownames_to_column("covariate") %>%
    ggplot(aes(covariate, pp)) +
    geom_bar(stat = "identity", fill = "lightblue4") +
    ylim(0, 2) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    labs(x = "", y = "-log10(adj. p)", title = "Association significance") +
    geom_hline(yintercept = -log10(0.05), colour = "red3")
), ncol = 5) -> p

@@

```{r, fig.width=25, fig.height=8}
cowplot::plot_grid(plotlist = list(
  cao_dis$plotSampleDistances(space = "expression.shifts", show.sample.size = T, method = "UMAP", title = "Condition"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$age %>% setNames(rownames(meta)), title = "Age"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$brain_bank %>% setNames(rownames(meta)), show.sample.size = T, title = "Brain bank"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$disease_duration %>% setNames(rownames(meta)), show.sample.size = T, title = "Disease duration"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$extract %>% setNames(rownames(meta)), show.sample.size = T, title = "Extract batch"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$flowcell %>% setNames(rownames(meta)), show.sample.size = T, title = "Flowcell"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$pmi %>% setNames(rownames(meta)), title = "PMI"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$sex %>% setNames(rownames(meta)), show.sample.size = T, title = "Sex"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$subtype %>% setNames(rownames(meta)), show.sample.size = T, title = "Subtype"),
  cao_dis$test.results$metadata.separation$pvalues %>%
    data.frame(p = .) %>%
    mutate(pp = -log10(p)) %>%
    tibble::rownames_to_column("covariate") %>%
    ggplot(aes(covariate, pp)) +
    geom_bar(stat = "identity", fill = "lightblue4") +
    ylim(0, 2) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    labs(x = "", y = "-log10(adj. p)", title = "Association significance") +
    geom_hline(yintercept = -log10(0.05), colour = "red3")
), ncol = 5) -> p

@@

@@

@@

# Extended Data Table 3

@@

@@

# Extended Data Table 13

@@

list(MSA = kegg.res, PD = kegg.res.pd) %>% 
  lapply(getElement, "result") %>% 
  lapply(dplyr::select, Description, GeneRatio, BgRatio, pvalue, p.adjust, ID, geneID) %>% 
  bind_rows(.id = "Disease") %>% 
  write.table("TableS13_KEGG.tsv", sep = "\t", row.names = F)
```

>>>>

# Extended Data Figure 4 & EDT3

## Load metadata

```{r}
meta <- read.delim("metadata.tsv", sep = "\t") %>%
  filter(sample %in% names(con$samples)) %>%
  tibble::column_to_rownames("sample") %>% 
  mutate(sample = rownames(.))
  # dplyr::select(-condition)
```

@@

```{r, fig.width=20, fig.height=8}
cowplot::plot_grid(plotlist = list(
  cao_pd$plotSampleDistances(space = "expression.shifts", show.sample.size = T, method = "UMAP", title = "Condition"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$age %>% setNames(rownames(meta)), title = "Age"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$brain_bank %>% setNames(rownames(meta)), show.sample.size = T, title = "Brain bank"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$extract %>% setNames(rownames(meta)), show.sample.size = T, title = "Extract batch"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$flowcell %>% setNames(rownames(meta)), show.sample.size = T, title = "Flowcell"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$pmi %>% setNames(rownames(meta)), title = "PMI"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$sex %>% setNames(rownames(meta)), show.sample.size = T, title = "Sex"),
  cao_pd$test.results$metadata.separation$padjust %>%
    data.frame(p = .) %>%
    mutate(pp = -log10(p)) %>%
    tibble::rownames_to_column("covariate") %>%
    ggplot(aes(covariate, pp)) +
    geom_bar(stat = "identity", fill = "lightblue4") +
    ylim(0, pmax(2, max(-log10(cao_pd$test.results$metadata.separation$padjust)))) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    labs(x = "", y = "-log10(adj. p)", title = "Association significance") +
    geom_hline(yintercept = -log10(0.05), colour = "red3")
), ncol = 4) -> p

@@

```{r, fig.width=25, fig.height=8}
cowplot::plot_grid(plotlist = list(
  cao_msa$plotSampleDistances(space = "expression.shifts", show.sample.size = T, method = "UMAP", title = "Condition"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$age %>% setNames(rownames(meta)), title = "Age"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$disease_duration %>% setNames(rownames(meta)), show.sample.size = T, title = "Disease duration"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$extract %>% setNames(rownames(meta)), show.sample.size = T, title = "Extract batch"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$flowcell %>% setNames(rownames(meta)), show.sample.size = T, title = "Flowcell"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$pmi %>% setNames(rownames(meta)), title = "PMI"),
    cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$sex %>% setNames(rownames(meta)), show.sample.size = T, title = "Sex"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$subtype %>% setNames(rownames(meta)), show.sample.size = T, title = "Subtype"),
  cao_msa$test.results$metadata.separation$padjust %>%
    data.frame(p = .) %>%
    mutate(pp = -log10(p)) %>%
    tibble::rownames_to_column("covariate") %>%
    ggplot(aes(covariate, pp)) +
    geom_bar(stat = "identity", fill = "lightblue4") +
    ylim(0, pmax(2, max(-log10(cao_pd$test.results$metadata.separation$padjust)))) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    labs(x = "", y = "-log10(adj. p)", title = "Association significance") +
    geom_hline(yintercept = -log10(0.05), colour = "red3")
), ncol = 5) -> p

@@

```{r, fig.width=25, fig.height=8}
cowplot::plot_grid(plotlist = list(
  cao_dis$plotSampleDistances(space = "expression.shifts", show.sample.size = T, method = "UMAP", title = "Condition"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$age %>% setNames(rownames(meta)), title = "Age"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$brain_bank %>% setNames(rownames(meta)), show.sample.size = T, title = "Brain bank"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$disease_duration %>% setNames(rownames(meta)), show.sample.size = T, title = "Disease duration"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$extract %>% setNames(rownames(meta)), show.sample.size = T, title = "Extract batch"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$flowcell %>% setNames(rownames(meta)), show.sample.size = T, title = "Flowcell"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$pmi %>% setNames(rownames(meta)), title = "PMI"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$sex %>% setNames(rownames(meta)), show.sample.size = T, title = "Sex"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$subtype %>% setNames(rownames(meta)), show.sample.size = T, title = "Subtype"),
  cao_dis$test.results$metadata.separation$padjust %>%
    data.frame(p = .) %>%
    mutate(pp = -log10(p)) %>%
    tibble::rownames_to_column("covariate") %>%
    ggplot(aes(covariate, pp)) +
    geom_bar(stat = "identity", fill = "lightblue4") +
    ylim(0, pmax(2, max(-log10(cao_pd$test.results$metadata.separation$padjust)))) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    labs(x = "", y = "-log10(adj. p)", title = "Association significance") +
    geom_hline(yintercept = -log10(0.05), colour = "red3")
), ncol = 5) -> p

@@

## Extended Data Table 3

```{r, eval = F}
list(cao_msa, cao_pd, cao_dis) %>% 
  lapply(getElement, "test.results") %>% 
  lget("metadata.separation") %>%
  lapply(\(x) x[-1]) %>%
  Map(\(dat, nn) bind_cols(dat) %>% data.frame() %>%  `rownames<-`(nn), dat = ., nn = lapply(., lapply, names) %>% sget(1)) %>% 
  setNames(c("MSA", "PD", "dis")) %>% 
  lapply(tibble::rownames_to_column, var = "covariate") %>% 
  bind_rows(.id = "comparison") %>% 
  mutate(comparison = scHelper::grepl.replace(comparison, c("MSA", "PD", "dis"), c("CTRL vs MSA", "CTRL vs PD", "PD vs MSA"))) %>% 
  write.table("/work/msa/05_tables/Table SX - Cacoa covariate analysis.tsv", sep = "\t", dec = ".", row.names = F)
```

@@

# Extended Data Figure 6

For panels a-d, see `scCODA.ipynb`.

Load data

```{r, fig.width=15, fig.height=8}
anno.major <- qread("anno_major.qs")
meta <- read.delim("metadata.tsv", sep = "\t")

anno.neurons <- qread("anno_neurons.qs")
```

## 6e

```{r}
# Minor final
anno.minor <- anno.major[!anno.major %in% c("PVMs", "Astrocytes", "Oligodendrocytes", "OPCs", "Microglia")] %>% 
  {factor(c(., anno.glia, anno.micro))} %>% 
  factor(levels = sort(levels(.)))

df <- data.frame(cid = names(anno.minor), anno = unname(anno.minor)) %>% 
  mutate(sample = strsplit(cid, "!!") %>% sget(1)) %>% 
  group_by(sample, anno) %>% 
  summarize(n = n()) %>% 
  mutate(prop = n / sum(n) * 100) %>% 
  ungroup() %>% 
  filter(anno %in% c("Inh. neurons", "Exc. neurons", "MSN")) %>% 
  mutate(age = meta$age[match(sample, meta$sample)]) %>% 
  mutate(age_bin = cut(age, breaks = c(0, 60, 70, 80, 90), labels = c("0-60", "60-70", "70-80", "80-90")),
         condition = grepl.replace(sample, c("CTRL", "MSA", "PD")))

df %>% 
  ggplot(aes(age_bin, prop)) + 
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.85)) + 
  theme_bw() +
  facet_wrap(~ anno) -> p

Statistics

```{r}
df %>% 
  group_by(anno) %>%
  rstatix::kruskal_test(prop ~ age_bin) %>% 
  mutate(padj = p.adjust(p, method = "BH"))
```

## 6f

```{r, fig.width=15, fig.height=8}
# Minor final
anno.minor <- anno.major[!anno.major %in% c("Inh. neurons", "MSN", "Exc. neurons", "PVMs", "Astrocytes", "Oligodendrocytes", "OPCs", "Microglia")] %>% 
  {factor(c(., anno.neurons, anno.glia, anno.micro))} %>% 
  factor(levels = sort(levels(.)))

df <- data.frame(cid = names(anno.minor), anno = unname(anno.minor)) %>% 
  mutate(sample = strsplit(cid, "!!") %>% sget(1)) %>% 
  group_by(sample, anno) %>% 
  summarize(n = n()) %>% 
  mutate(prop = n / sum(n) * 100) %>% 
  ungroup() %>% 
  filter(anno %in% levels(anno.micro)) %>% 
  mutate(age = meta$age[match(sample, meta$sample)]) %>% 
  mutate(age_bin = cut(age, breaks = c(0, 60, 70, 80, 90), labels = c("0-60", "60-70", "70-80", "80-90")),
         condition = grepl.replace(sample, c("CTRL", "MSA", "PD")))

df %>% 
  ggplot(aes(age_bin, prop)) + 
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.85)) + 
  theme_bw() +
  facet_wrap(~ anno) -> p

Statistics

```{r}
df %>% 
  group_by(anno) %>%
  rstatix::kruskal_test(prop ~ age_bin) %>% 
  mutate(padj = p.adjust(p, method = "BH"))
```

@@

@@

# Extended Data Table 11

@@

# Extended Data Table 14

@@

list(MSA = kegg.res, PD = kegg.res.pd) %>% 
  lapply(getElement, "result") %>% 
  lapply(dplyr::select, Description, GeneRatio, BgRatio, pvalue, p.adjust, ID, geneID) %>% 
  bind_rows(.id = "Disease") %>% 
  write.table("TableS14_KEGG.tsv", sep = "\t", row.names = F)
```
